Testing Quasar Unification with Clumpy Wind Models

### 2015-11-10 paper.tex
fixed edits

Perhaps the clearest evidence of outflows in AGN is  
the blueshifted ($\sim 0.1c$) broad absorption lines (BALs) in the 
ultraviolet seen in approximately $20\%$ of quasars
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within the ADW paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}, although there is also evidence
for polar BAL outflows \citep{zhou2006}.
% although this geometry is contradicted
% by more recent studies of radio spectral indices \citep{dipompeo2012}.
ADWs offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.

@@

The UV spectrum is characterised by strong BAL 
profiles at high inclinations ($> 70^\circ$). 
This highlights the first success of our model: 
clumping means the correct ionization state 
is maintained in the presence of strong X-rays, 
allowing large resonance line opacities. 
At the highest inclinations, the 
cooler, lower ionization material at the base of the wind
starts to intersect the line of sight. This produces 
multiple absorption lines in lower ionization species such as \mg,
\al\ and Fe~\textsc{ii}. The potential links to LoBALQSOs and 
FeLoBALQSOs are discussed in section 2.4.

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At lower inclinations, the strongest absorption occurs at the red edge,
whereas at high inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008, Filiz Ak et al. 2014).
In addition, the line profile shape is strongly dependent 
on the density, ionization and velocity 
profiles intersected by the line of sight. Thus, small tweaks of the velocity
law and angular distributions of streamlines can dramatically alter
the shape of the line.

@@

BALQSOs and quasars generally possess very similar emission line properties 
\citep[e.g.][]{weymann1991,reichard2003}. To demonstrate this, we show equivalent width
mean values and standard devaitions from REF in table 3. Clearly, the variation
with inclination in our models is far greater than than the variation 
between quasar to BALQSO samples.  
This presents a challenge to our model, as well as the geometric unification picture in general.
Obviously,  one possible solution to this problem is hypothesize a more isotropic distribution for the emergent condition than predicted by a ``classical'' disc. 
General relativistic effects -- specifically, light bending
and relativistic beaming -- can cause 
the accretion disc SED to become more isotropic \citep[e.g.][]{zhang1997,munozdarias2013}. 
However this effect is small in the UV and optical wavelength regimes, and the 
disc is still anisotropic.

Reprocessing by an extended outflow may also cause a more isotropic
continuum to emerge. Hints that light scattered off a spatially extended 
wind may contribute signiciantly to the emergent continuum come from radiative transfer s
simulations \citep{simproga2012} and microlensing observations \citep{sluse2015}.
However, neither of these examples have sufficient reprocessing efficiencies to 
compensate for the disc anisotropy in this case.
An alternative explanation, suggested by \cite{risaliti2011}, is that
the BLR has the same angular distribution of emission as the accretion disc.
They find that EW distributions in quasars are consistent with anisotropic emission 
from optically thick, disc-like structures for {\em both} the continuum source and BLR. 
If this is the case, it has a dramatic affect on the intrinsic
BAL fraction inferred from flux-limited samples \citep{goodrich1997,krolikvoit1998}.

The final possibility is that BALQSOs are not viewed equatorially. 
This possibility is supported by measurements of the radio spectral indices
of radio-loud BALQSOs (REF) and attempts to explain the polarisation dichotomy 
of type 1 and 2 AGN (REF). This would imply that type 1 quasars and BALQSOs are generally 
viewed from a fairly narrow range of angles ($0$-$50^\circ$, and could also mean 
the BALQSO fraction requires {\em both} evolutionary and geometric explanations. 
There is also clear conflict between this explanation and some
polarisation results \citep{goodrich1995, cohen1995}(BROTHERTON).

@@

At certain sightlines, the synthetic spectra exhibit blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs, and we even see absorption in Fe~\textsc{ii}
at the highest inclinations. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in figure~\ref{fig:lobal} for a range
of BALQSO viewing angles. We find that ionization stratification
of the wind causes lower ionization material to have a smaller covering factor, 
as demonstrated by figures~\ref{fig:wind} and \ref{fig:lobal}.
This confirms the behaviour expected from a unification model such as Elvis (2000). 
LoBALs are only present at viewing angles close to edge-on ($i>75^\circ$),
as predicted by polarisation results \citep{brotherton1997}.
As observed in a BALQSO sample by \cite{filizak2014}, we find that
BAL troughs are wider and deeper when lower ionization species are present,
and high ionization lines have higher blue-edge velocities than the 
lower ionization species.
There is also a correlation between the strength of LoBAL features
and the amount of continuum attenuation at that sightline, particularly
blueward of the Lyman edge as the low ionization base 
intersects the line-of-sight. 
A model such as this therefore predicts that LoBALQSOs and FeLoBALQSOs 
have stronger Lyman edge absorption and 
are more Compton-thick than HiBALQSOs and Type 1 quasars.
An edge-on scenario also offers a potential explanation for the rarity of LoBAL and
FeLoBAL quasars, due to a foreshortened and attenuated continuum, 
although, as noted in section~\ref{discsed}, BAL fraction 
inferences are fraught with complex selection effects.

@@

The work of JHM, SWM, NSH and CK is supported by the 
Science and Technology Facilities Council (STFC), 
via two studentships and a consolidated grant, respectively.
We would like to thank Omer Blaes, Ivan Hubeny and Shane Davis for their
assistance with \agn. We would also like to thank Daniel Proga, 
Daniel Capellupo, Sam Connolly and Dirk Grupe for useful discussions.
Simulations were conducted using \py\ version 79c, and made use of
the IRIDIS High Performance Computing Facility at the University of Southampton.
Figures were produced using {\tt matplotlib} \citep{matplotlib}.

>>>>

Perhaps the clearest evidence of outflows in AGN is  
the blueshifted ($\sim 0.1c$) broad absorption lines (BALs) in the 
ultraviolet seen in approximately $20\%$ of quasars
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within the ADW paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}, although there is also evidence
for polar BAL outflows  in radio-loud (RL) sources \citep{zhou2006}.
% although this geometry is contradicted
% by more recent studies of radio spectral indices \citep{dipompeo2012}.
ADWs offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.

@@

The UV spectrum is characterised by strong BAL 
profiles at high inclinations ($> 70^\circ$). 
This highlights the first success of our model: 
clumping means the correct ionization state 
is maintained in the presence of strong X-rays, 
allowing large resonance line opacities. 
At the highest inclinations, the 
cooler, low ionization material at the base of the wind
starts to intersect the line of sight. This produces 
multiple absorption lines in species such as \mg,
\al\ and Fe~\textsc{ii}. The potential links to LoBALQSOs and 
FeLoBALQSOs are discussed in section 2.4.

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At the lowest inclination BAL sightlines, the strongest absorption occurs at the red edge,
whereas at higher inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008, Filiz Ak et al. 2014).
In addition, the line profile shape is strongly dependent 
on the density, ionization and velocity 
profiles intersected by the line of sight. Thus, small tweaks of the velocity
law and angular distributions of streamlines can dramatically alter
the shape of the line.

@@

BALQSOs and quasars generally possess very similar emission line properties 
\citep[e.g.][]{weymann1991,reichard2003}. To demonstrate this, we show equivalent width
mean values and standard devaitions from REF in table 3. Clearly, the variation
with inclination in our models is far greater than than the variation 
between, and standard deviation within, quasar and BALQSO samples.  
This presents a challenge to our model, as well as the geometric unification picture in general.
One obvious potential solution to this problem is to hypothesize a more isotropic distribution 
for the emergent condition than predicted by a classical thin disc. 
General relativistic effects -- specifically, light bending
and relativistic beaming -- can cause 
the accretion disc SED to become more isotropic \citep[e.g.][]{zhang1997,munozdarias2013}. 
However, we have verified using AGNSPEC (REFs to Hubeny)
that this effect is small in the UV and optical wavelength regimes, 
and the disc is still anisotropic.

Reprocessing by an extended outflow may also cause a more isotropic
continuum to emerge. Hints that light scattered off a spatially extended 
wind may contribute signiciantly to the emergent continuum come from radiative transfer s
simulations \citep{simproga2012} and microlensing observations \citep{sluse2015}.
However, neither of these examples have sufficient reprocessing efficiencies to 
compensate for the disc anisotropy in this case.
An alternative explanation is that the BLR has the same angular distribution 
of emission as the accretion disc.
Indeed, \cite{risaliti2011} find that EW distributions in quasars are consistent with anisotropic emission 
from optically thick, disc-like structures for {\em both} the continuum source and BLR. 
If this is the case, it has a dramatic affect on the intrinsic
BAL fraction inferred from flux-limited samples \citep{goodrich1997,krolikvoit1998}.

It is also possible that the equatorial paradigm invoked from early polarisation
studies \citep{goodrich1995, cohen1995,brotherton2006} is an over-simplification, or is merely incorrect. 
High brightness temperatures in some RL BALQSOs imply polar outflows \citep{zhou2006} 
and \cite{bruni2012} find that
RL BALQSOs possess similar radio spectral indices to normal RL quasars, 
suggestive of comparable inclinations. 
In addition, \cite{marin2013} find a bending angle of $\sim45^\circ$ is required to 
explain the polarisation dichotomy of type 1 and 2 AGN using an Elvis-type wind 
model (Elvis 2000). It is therefore possible 
that type 1 quasars and BALQSOs are generally viewed from a fairly narrow range of angles 
($\sim0$-$45^\circ$), or that {\em both} evolutionary and geometric explanations are required.
We suggest that future modelling should include predictions of polarisation signatures
from a detailed radiative transfer simulation, allowing direct comparison with spectropolarimetry
of BALQSOs.

% The final possibility is that BALQSOs are not viewed equatorially. 
% This possibility is supported by measurements of the radio spectral indices
% of radio-loud BALQSOs (REF) and attempts to explain the polarisation dichotomy 
% of type 1 and 2 AGN (REF). This would imply that type 1 quasars and BALQSOs are generally 
% viewed from a fairly narrow range of angles ($\sim0$-$50^\circ$), and could also mean 
% the BALQSO fraction requires {\em both} evolutionary and geometric explanations. 
% There is also clear conflict between this explanation and some
% polarisation results \citep{goodrich1995, cohen1995}(BROTHERTON).
% We suggest that future modelling should include predictions of polarisation signatures
% from a detailed radiative transfer simulation, allowing

@@

At certain sightlines, the synthetic spectra exhibit blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs, and we even see absorption in Fe~\textsc{ii}
at the highest inclinations. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in figure~\ref{fig:lobal} for a range
of BALQSO viewing angles. We find that ionization stratification
of the wind causes lower ionization material to have a smaller covering factor, 
as demonstrated by figures~\ref{fig:wind} and \ref{fig:lobal}.
This confirms the behaviour expected from a unification model such as Elvis (2000). 
LoBALs are only present at viewing angles close to edge-on ($i>75^\circ$),
as predicted by polarisation results \citep{brotherton1997}.
As observed in a BALQSO sample by \cite{filizak2014}, we find that
BAL troughs are wider and deeper when low ionization absorption features are present,
and high ionization lines have higher blue-edge velocities than the 
low ionization species.
There is also a correlation between the strength of LoBAL features
and the amount of continuum attenuation at that sightline, particularly
blueward of the Lyman edge as the low ionization base 
intersects the line-of-sight. 
A model such as this therefore predicts that LoBALQSOs and FeLoBALQSOs 
have stronger Lyman edge absorption and 
are more Compton-thick than HiBALQSOs and Type 1 quasars.
An edge-on scenario also offers a potential explanation for the rarity of LoBAL and
FeLoBAL quasars, due to a foreshortened and attenuated continuum, 
although, as noted in section~\ref{discsed}, BAL fraction 
inferences are fraught with complex selection effects.

@@

The work of JHM, SWM, NSH and CKL is supported by the
Science and Technology Facilities Council (STFC),
via two studentships and a consolidated grant, respectively.
CK also acknowledges a Leverhulme fellowship.
We would like to thank Omer Blaes, Ivan Hubeny and Shane Davis for their
assistance with \agn. We are grateful to Mike Brotherton, Mike DiPompeo,
Sebastien Hoenig and Frederic Marin for helpful correspondence regarding
polarisation measurements and orientation indicators.
We would also like to thank Daniel Proga, Daniel Capellupo, Sam Connolly and
Dirk Grupe for useful discussions.  Simulations were conducted using \py\ version 80,
and made use of the IRIDIS High Performance Computing Facility at the
University of Southampton. Figures were produced using the {\tt matplotlib} plotting library
\citep{matplotlib}.
